---
title: "Visual Identity Guide — Biostatistika (MB120P163)"
subtitle: "Colour, typography, and figure conventions for all course materials"
author: "Ondřej Mottl"
lang: en
date: 2026/03/26
date-modified: last-modified
---

```{r}
#| label: setup
#| include: false

library(here)
library(jsonlite)
library(ggplot2)
library(tibble)
library(purrr)

here::i_am("brand_guidelines.qmd")

colors_data <-
  jsonlite::fromJSON(
    here::here("quarto/colors.json")
  )
fonts_data <-
  jsonlite::fromJSON(
    here::here("quarto/fonts.json")
  )

primary <- colors_data$primary

# Named colour constants reused throughout this document
clr_data      <- primary[["grey_olive"]]
clr_model     <- primary[["indigo_velvet"]]
clr_model2    <- primary[["amethyst"]]
clr_highlight <- primary[["orange"]]
clr_bg        <- primary[["parchment"]]
clr_text      <- primary[["graphite"]]
```

## Purpose

This document defines the visual identity system for the course **Biostatistika (MB120P163)**
at the Faculty of Science, Charles University.

The goal is to ensure:

- conceptual clarity (data vs model vs interpretation)
- visual consistency across all outputs
- low cognitive load for students

### Language requirement

All student-facing content — axis labels, callout headings, interaction prompts, figure
captions, slide text, and annotations — must be in **Czech**. This is a hard requirement,
not a preference. The only exceptions are:

- R function names, package names, and error messages (remain in English)
- Internal technical documentation and README files (English — read by collaborators, not students)
- Inline R code spans and code blocks

---

## Core Design Philosophy

The visual system encodes **epistemic roles**:

- **Data (observations)** → neutral, grey
- **Model (inference)** → purple
- **Interpretation / emphasis** → orange

Colour is not decorative. It conveys meaning.

### Connection to the pedagogical sequence

The colour order mirrors the course's canonical teaching sequence:

| Step | What happens | Colour |
| ---- | ------------ | ------ |
| 1 | Biological question is posed | — (no data yet) |
| 2 | Data are shown | grey |
| 3 | Students predict / interpret | — (question slide) |
| 4 | Statistical result is revealed | purple |
| 5 | Result is interpreted biologically | orange |

A figure that uses these colours in the correct semantic roles therefore *mirrors the
analytical workflow students are being trained to follow*. This is the primary justification
for treating colour as meaning rather than decoration.

---

## Colour Palette

The canonical palette is defined in [`quarto/colors.json`](quarto/colors.json).
Every lecture repo downloads this file at render time.

### Base colours

- `parchment` (`r primary[["parchment"]]`) — background
- `graphite` (`r primary[["graphite"]]`) — primary text

### Neutral / Data layer

- `grey_olive` (`r primary[["grey_olive"]]`) — data points, gridlines

### Model layer (primary accent)

- `indigo_velvet` (`r primary[["indigo_velvet"]]`) — main model
- `amethyst` (`r primary[["amethyst"]]`) — secondary model / comparison

### Highlight layer

- `orange` (`r primary[["orange"]]`) — emphasis, key results

### Live colour swatches

```{r}
#| label: colour-swatches
#| echo: false
#| fig-width: 8
#| fig-height: 1.8
#| fig-alt: "Colour swatches for the course brand palette"

swatch_df <-
  tibble::tibble(
    name = names(primary),
    hex  = unlist(primary),
    x    = seq_along(primary)
  ) |>
  dplyr::mutate(
    # Choose readable label colour based on luminance
    label_col = dplyr::if_else(
      name %in% c("graphite"),
      primary[["parchment"]],
      primary[["graphite"]]
    )
  )

ggplot2::ggplot(
  swatch_df,
  ggplot2::aes(x = x, y = 1)
) +
  ggplot2::geom_tile(
    ggplot2::aes(fill = hex),
    width = 0.95,
    height = 0.7
  ) +
  ggplot2::geom_text(
    ggplot2::aes(label = name, colour = label_col),
    y = 1.08,
    size = 3,
    fontface = "bold",
    vjust = 0
  ) +
  ggplot2::geom_text(
    ggplot2::aes(label = hex, colour = label_col),
    y = 0.92,
    size = 2.6,
    vjust = 1
  ) +
  ggplot2::scale_fill_identity() +
  ggplot2::scale_colour_identity() +
  ggplot2::scale_x_continuous(expand = ggplot2::expansion(mult = 0.05)) +
  ggplot2::scale_y_continuous(limits = c(0.6, 1.4)) +
  ggplot2::labs(x = NULL, y = NULL, title = "Course colour palette") +
  ggplot2::theme_void() +
  ggplot2::theme(
    plot.title = ggplot2::element_text(
      size = 11,
      face = "bold",
      colour = primary[["graphite"]],
      margin = ggplot2::margin(b = 6)
    ),
    plot.background = ggplot2::element_rect(
      fill = primary[["parchment"]],
      colour = NA
    )
  )
```

---

## Semantic Mapping (MANDATORY)

| Role              | Colour name    | Hex value                       | Usage                                  |
| ----------------- | -------------- | ------------------------------- | -------------------------------------- |
| Data              | `grey_olive`   | `r primary[["grey_olive"]]`     | raw observations, background structure |
| Model (primary)   | `indigo_velvet`| `r primary[["indigo_velvet"]]`  | main fitted curve, main inference      |
| Model (secondary) | `amethyst`     | `r primary[["amethyst"]]`       | alternative models, comparisons        |
| Highlight         | `orange`       | `r primary[["orange"]]`         | key insight, important value, warning  |

Rules:

- Grey is **always** data
- Purple family is **always** modelling
- Orange is used **sparingly** for emphasis

---

## Usage Rules

### General

- Use maximum **3 colours per figure**
- Maintain strong contrast hierarchy
- Prefer restraint over expressiveness

### Data (points)

- Colour: `grey_olive`
- Alpha: 0.2–0.4
- Size: small

### Models (lines)

- Primary model: `indigo_velvet`
- Secondary model: `amethyst`
- Line width: slightly thicker than default

### Highlights

Use `orange` only for:

- key datapoints
- annotated regions
- important labels

Do **NOT**:

- colour entire plots orange
- use orange for decorative purposes

### Uncertainty

- Use transparency of purple (`alpha = 0.15`)
- Or lighter amethyst
- Avoid introducing new colours

---

## Typography

Font definitions are in [`quarto/fonts.json`](quarto/fonts.json).

### Primary (sans-serif)

- Body: **`r fonts_data$body`** (weights 400, 500, 700)
- Headings: **`r fonts_data$heading`** (weights 400, 600, 700)

### Monospace

- **`r fonts_data$monospace`**; fallback: Fira Code

Usage:

- Monospace for code, formulas, filenames
- Optional: purple accent (`indigo_velvet`) for model formulas

---

## Layout Principles

- Left-aligned layouts
- Generous whitespace
- One main visual per slide

Optional structural motif:

- Thin vertical purple line for consistent framing

---

## Figure Design Pattern

Standard layer order:

1. Grey points (data)
2. Purple curve (model)
3. Optional secondary curve (amethyst)
4. Orange highlight (if needed)

This order reflects the analytical workflow (§2.1).

### Biological framing requirement

Every figure in a student-facing context must be anchored in a concrete biological or
ecological question.

Required:

- **Caption or slide narrative** states the biological question the figure addresses
  (e.g. *"Závisí délka křídla na druhu tučňáka?"*)
- **Axis labels** use ecologically meaningful Czech names with units — never raw column names
- **Annotations** (if orange) explain the biological significance, not just the statistical value

A figure is complete only when a student could answer "what biological question does this
address?" from the figure and its caption alone.

---

## ggplot2 Implementation

### Named colour constants

Define these at the top of every setup chunk and reuse throughout:

```r
clr_data      <- "#8A8A8A"   # grey_olive    — raw observations
clr_model     <- "#5D2890"   # indigo_velvet — primary model
clr_model2    <- "#86579E"   # amethyst      — secondary model / comparison
clr_highlight <- "#F3A712"   # orange        — emphasis / key result
clr_bg        <- "#F4F1EC"   # parchment     — plot background
```

### Standard course ggplot2 theme (`theme_biostat`)

Every figure must end with `+ theme_biostat()`. Define this function once in each
lecture's `R/set_r_theme.R` (generated automatically by `R/generate_theme.R`).

Reference implementation:

```r
theme_biostat <- function() {
  ggplot2::theme_minimal(base_family = "Source Sans 3") +
    ggplot2::theme(
      plot.background   = ggplot2::element_rect(fill = "#F4F1EC", colour = NA),
      panel.background  = ggplot2::element_rect(fill = "#F4F1EC", colour = NA),
      panel.grid.major  = ggplot2::element_line(colour = "#E0DDD8"),
      panel.grid.minor  = ggplot2::element_blank(),
      axis.text         = ggplot2::element_text(colour = "#2E2E2E"),
      axis.title        = ggplot2::element_text(colour = "#2E2E2E"),
      plot.title        = ggplot2::element_text(colour = "#2E2E2E", face = "bold"),
      legend.background = ggplot2::element_rect(fill = "#F4F1EC", colour = NA)
    )
}
```

### Example: scatter plot (data only) {.unnumbered}

```{r}
#| label: example-scatter-data
#| echo: true
#| fig-width: 6
#| fig-height: 4
#| fig-alt: "Example scatter plot showing grey data points on a parchment background"

# Synthetic data for illustration
set.seed(1)
n <- 80
example_df <-
  tibble::tibble(
    x = rnorm(n, mean = 20, sd = 5),
    y = 0.8 * x + rnorm(n, sd = 3)
  )

ggplot2::ggplot(
  example_df,
  ggplot2::aes(x = x, y = y)
) +
  ggplot2::geom_point(
    colour = clr_data,
    alpha  = 0.35,
    size   = 2
  ) +
  ggplot2::labs(
    x     = "Predictor variable (units)",
    y     = "Response variable (units)",
    title = "Data only — grey points on parchment"
  ) +
  ggplot2::theme_minimal(base_family = "sans") +
  ggplot2::theme(
    plot.background  = ggplot2::element_rect(fill = clr_bg, colour = NA),
    panel.background = ggplot2::element_rect(fill = clr_bg, colour = NA),
    panel.grid.major = ggplot2::element_line(colour = "#E0DDD8"),
    panel.grid.minor = ggplot2::element_blank()
  )
```

### Example: scatter + fitted model {.unnumbered}

```{r}
#| label: example-scatter-model
#| echo: true
#| fig-width: 6
#| fig-height: 4
#| fig-alt: "Example scatter plot showing grey data points and a purple fitted line"

ggplot2::ggplot(
  example_df,
  ggplot2::aes(x = x, y = y)
) +
  ggplot2::geom_point(
    colour = clr_data,
    alpha  = 0.35,
    size   = 2
  ) +
  ggplot2::geom_smooth(
    method    = "lm",
    colour    = clr_model,
    linewidth = 1.2,
    se        = FALSE,
    formula   = y ~ x
  ) +
  ggplot2::labs(
    x     = "Predictor variable (units)",
    y     = "Response variable (units)",
    title = "Data (grey) + model (purple)"
  ) +
  ggplot2::theme_minimal(base_family = "sans") +
  ggplot2::theme(
    plot.background  = ggplot2::element_rect(fill = clr_bg, colour = NA),
    panel.background = ggplot2::element_rect(fill = clr_bg, colour = NA),
    panel.grid.major = ggplot2::element_line(colour = "#E0DDD8"),
    panel.grid.minor = ggplot2::element_blank()
  )
```

---

## Consistency Rules

- Do not reassign colour meanings
- Do not introduce new hues without justification
- Maintain consistency across lectures and materials

---

## Guiding Principle

> Colour encodes epistemic status, not category.

- grey → observed
- purple → modelled
- orange → interpreted

---

## Default Decision Rules

Use the following defaults unless there is a strong reason to deviate:

1. Map data to grey (`grey_olive`)
2. Map the primary model to `indigo_velvet`
3. Use `amethyst` only for secondary comparison
4. Use `orange` only for deliberate emphasis
5. Limit each figure to the smallest viable palette
6. Prefer minimalism, consistency, and semantic clarity

Safe default (for most figures): grey points + purple model + no accent colour.

---

## Future Extensions

Possible controlled extensions:

- line types for additional models (instead of adding colours)
- transparency gradients for density
- annotations instead of additional colours

Avoid expanding the palette unless strictly necessary.

---

## Presentations (Quarto RevealJS)

### Slide background and theme

- Background colour: `parchment` (`r clr_bg`)
- Primary text: `graphite` (`r clr_text`)
- Code highlight style: `github-light` (consistent with parchment background)

### Typography on slides

- Headings: Source Sans 3 or Inter, bold
- Body: same sans-serif family at regular weight
- Code: JetBrains Mono or Fira Code
- Do NOT use more than one heading weight per slide

### Colour use in slide text and callouts

- Use `indigo_velvet` for highlighted model formulas or key statistical terms — apply sparingly
- Use `orange` for "question" labels, key-result callout borders, or "important" boxes
- Do NOT colour regular prose text for decoration
- Callout with left purple border = supplementary explanation; orange border = call to attention

### Figure density

- One figure per slide is the default
- Two-column layout is allowed for: side-by-side model comparisons, before/after reveals,
  data vs model overlays
- Never place more than two distinct figures on a single slide

### Progressive disclosure

- Reveal layers in this order: data (grey) → model (purple) → interpretation (orange callout)
- Use RevealJS fragments or incremental list options
- Never show a model result before posing a prediction question to the audience

### Interaction slides

- Mark interaction slides with a thick `indigo_velvet` left border or a distinct callout block
- The word "Otázka" or "Diskuze" can be accented in `orange`
- Keep ALL interaction slides visually consistent within a single lecture

### Code display on slides

- Keep visible code chunks to ≤ 8 lines; scrolling is acceptable for demos
- Code output inherits the parchment background — do not override with white or grey
- Student-visible code uses base R only (no tidyverse prefixes)

---

## Skripta (Learning Materials — Quarto HTML / Typst PDF)

### HTML theme

- Base Quarto theme: `flatly` + custom `skripta_theme.scss` (generated from JSON)
- TOC: always shown (`toc: true`, `toc-depth: 3`), left sidebar in HTML
- Section numbering: always enabled (`number-sections: true`)

### Callout blocks

Use Quarto callouts for exactly four purposes:

| Callout type | Purpose | Accent colour |
| ------------ | ------- | ------------- |
| `note` | Supplementary explanation, extra detail | indigo_velvet (SCSS customised) |
| `tip` | Optional technical context that is not needed for the main argument | grey_olive |
| `important` | Key concept or definition the student must retain | indigo_velvet |
| `warning` | Common mistake or statistical misconception | orange |

Use `tip` only with the Czech heading **Doplňující** (optionally followed by a short descriptor). For a long optional code example, place a brief `tip` callout before the code and fold the code in HTML; the code remains in the normal page flow in PDF. A short self-contained technical example may instead be placed directly in a collapsible `tip` callout, which remains expanded in PDF. Avoid `caution` and do not invent further callout types.

### Colour in prose

- Do not colour prose text
- **Bold** is the only allowed inline emphasis for key terms appearing for the first time
- *Italic* is reserved for species names and foreign-language terms — not for general emphasis
- Code inline spans are used for all R objects, function names, variable names, and file paths

### Figures in skripta

- Semantic colour mapping applies identically to skripta figures
- Every figure requires a caption that **interprets** the figure — not just describes it
- Default width: `out.width = "80%"` for HTML; `out.width = "100%"` for Typst/PDF
- Always include `fig-alt` alt text for accessibility

### Interactive HTML components

#### Glossary tooltips

- Tooltip background: `#FFFFFF` (white) — must contrast against the parchment (`#F4F1EC`) page background
- Border: `1px solid rgba(93, 40, 144, 0.25)` — subtle indigo_velvet tint
- Shadow: `0 4px 12px rgba(0, 0, 0, 0.12)` — lifts the tooltip visually off the page
- The CSS is emitted by `nastaveni-slovnik` chunk; the `glossary_style()` call must use `def_bg = "#FFFFFF"`

#### Tabsets (`.panel-tabset`)

Tab button states use existing colour tokens to give clear click affordance against the parchment background:

| State | Background | Text | Border |
| ----- | ---------- | ---- | ------ |
| Inactive | `light_gray` (`#F0EDE8`) | `graphite` | `rgba(0,0,0,0.12)` |
| Hover | `amethyst` (`#86579E`) | white | `amethyst` |
| Active | `indigo_velvet` (`#5D2890`) | white | `indigo_velvet` |

This styling is generated automatically by `generate_skripta_html_theme.R` and must not be overridden manually.

### Code blocks

- Student-visible chunks (`echo: true`): base R, no prefixes for preloaded packages
- Hidden setup chunks (`echo: false`): tidyverse with full namespace (`pkg::fn()`)
- Code folding is acceptable only for lengthy raw data preparation

---

## R Visualisations — Extended Reference

### Model uncertainty bands

Show uncertainty via transparent fill using the model colour — do NOT introduce a new colour:

```r
ggplot2::geom_ribbon(
  ggplot2::aes(ymin = lower, ymax = upper),
  fill  = clr_model,
  alpha = 0.15
)
```

### Two-model comparison

Use a thinner, dashed line for the secondary model to reinforce the visual hierarchy:

```r
# Primary model
ggplot2::geom_smooth(colour = clr_model,  linewidth = 1.2, se = FALSE)
# Secondary model
ggplot2::geom_smooth(colour = clr_model2, linewidth = 1.0, se = FALSE, linetype = "dashed")
```

### Boxplots and violin plots

Data points underneath a transparent model-coloured box:

```r
ggplot2::geom_jitter(colour = clr_data,  alpha = 0.3,  width = 0.15, size = 1.5) +
ggplot2::geom_boxplot(colour = clr_model, fill = NA, linewidth = 0.8, outlier.shape = NA)
```

### Categorical group colour

When colour **must** encode a grouping variable (e.g. species, treatment):

- Use a distinct palette that does NOT overlap with the semantic palette
- Never reuse `clr_model` or `clr_highlight` as a group colour
- Document the override with a comment
- Suggested accessible palette for up to 4 groups:

```r
clr_skupiny <- c(
  "skupina_A" = "#4E9A8A",   # teal
  "skupina_B" = "#C2613A",   # terracotta
  "skupina_C" = "#2E6CA4",   # steel blue
  "skupina_D" = "#7D7D7D"    # dark grey
)
```

### Highlights and annotations

```r
# Highlighted individual point
ggplot2::geom_point(
  data   = data_highlighted,
  colour = clr_highlight,
  size   = 4
)

# Annotation label
ggplot2::annotate(
  geom     = "text",
  x        = x_pos,
  y        = y_pos,
  label    = "Klíčový výsledek",
  colour   = clr_highlight,
  fontface = "bold",
  size     = 4
)
```

### Figure dimensions by output context

| Context | `fig.width` | `fig.height` | `out.width` |
| ------- | ----------- | ------------ | ----------- |
| Presentation slide | 7 | 5 | `"100%"` |
| Skripta HTML | 7 | 5 | `"80%"` |
| Skripta PDF (Typst) | 6 | 4 | `"100%"` |

### Axis labels and Czech language

- Always use Czech axis labels in student-visible plots; set with `ggplot2::labs()`
- Never leave default English column-name axis labels in student-facing output
- Include units in parentheses where relevant: `"Délka křídla (mm)"`
- Annotation text passed to `ggplot2::annotate()` must be Czech
- Legend labels and guide titles must be Czech

See §1.1 for the global Czech language requirement.

---

## GitHub / Repository Identity

### README files

- Language: English (technical documentation is public-facing)
- Content: state what the repo contains, how to render locally, and who the audience is
- Keep concise; do not include personal context or course history

### Organisation profile (`.github` repo)

- The `README.md` in the `.github` repo is the public face of the organisation
- Colour references should use hex values where relevant
- Tone: informative and neutral

### Release assets and file naming

- Rendered outputs use predictable, fixed names: `presentation.html`, `skripta.html`, `skripta_raw.pdf`
- File names must not contain spaces or special characters (including Czech diacritics)
- Do not change naming conventions without updating the corresponding GitHub Actions workflow
