---
title: "Visual Identity Guide — Biostatistika (MB120P163)"
subtitle: "Colour, typography, and figure conventions for all course materials"
author: "Ondřej Mottl"
lang: en
date: 2026/03/26
date-modified: last-modified
---

```{r}
#| label: setup
#| include: false

library(here)
library(jsonlite)
library(ggplot2)
library(tibble)
library(purrr)

here::i_am("brand_guidelines.qmd")

colors_data <-
  jsonlite::fromJSON(
    here::here("quarto/colors.json")
  )
fonts_data <-
  jsonlite::fromJSON(
    here::here("quarto/fonts.json")
  )

primary <- colors_data$primary

# Named colour constants reused throughout this document
clr_data      <- primary[["grey_olive"]]
clr_model     <- primary[["indigo_velvet"]]
clr_model2    <- primary[["amethyst"]]
clr_highlight <- primary[["orange"]]
clr_bg        <- primary[["parchment"]]
clr_text      <- primary[["graphite"]]

theme_biostat <- function(
  base_size = 11,
  base_family = "Source Sans 3"
) {
  output_background <-
    if (knitr::is_html_output()) clr_bg else "white"

  ggplot2::theme_bw(
    base_size = base_size,
    base_family = base_family
  ) +
    ggplot2::theme(
      text = ggplot2::element_text(colour = clr_text),
      panel.border = ggplot2::element_blank(),
      axis.line = ggplot2::element_line(colour = clr_text, linewidth = 0.4),
      panel.background = ggplot2::element_rect(fill = "white", colour = NA),
      plot.background = ggplot2::element_rect(fill = output_background, colour = NA),
      legend.background = ggplot2::element_rect(
        fill = output_background,
        colour = NA
      ),
      legend.key = ggplot2::element_rect(fill = output_background, colour = NA),
      panel.grid.minor = ggplot2::element_blank(),
      legend.position = "bottom"
    )
}
```

## Purpose

This document defines the visual identity system for the course **Biostatistika (MB120P163)**
at the Faculty of Science, Charles University.

The goal is to ensure:

- conceptual clarity (data vs model vs interpretation)
- visual consistency across all outputs
- low cognitive load for students

### Language requirement

All student-facing content — axis labels, callout headings, interaction prompts, figure
captions, slide text, and annotations — must be in **Czech**. This is a hard requirement,
not a preference. The only exceptions are:

- R function names, package names, and error messages (remain in English)
- Internal technical documentation and README files (English — read by collaborators, not students)
- Inline R code spans and code blocks

---

## Core Design Philosophy

The visual system encodes **epistemic roles**:

- **Data (observations)** → neutral, grey
- **Model (inference)** → purple
- **Interpretation / emphasis** → orange

Colour is not decorative. It conveys meaning.

### Connection to the pedagogical sequence

The colour order mirrors the course's canonical teaching sequence:

| Step | What happens | Colour |
| ---- | ------------ | ------ |
| 1 | Biological question is posed | — (no data yet) |
| 2 | Data are shown | grey |
| 3 | Students predict / interpret | — (question slide) |
| 4 | Statistical result is revealed | purple |
| 5 | Result is interpreted biologically | orange |

A figure that uses these colours in the correct semantic roles therefore *mirrors the
analytical workflow students are being trained to follow*. This is the primary justification
for treating colour as meaning rather than decoration.

---

## Colour Palette

The canonical palette is defined in [`quarto/colors.json`](quarto/colors.json).
Every lecture repo downloads this file at render time.

### Base colours

- `parchment` (`r primary[["parchment"]]`) — background
- `graphite` (`r primary[["graphite"]]`) — primary text

### Neutral / Data layer

- `grey_olive` (`r primary[["grey_olive"]]`) — data points, gridlines

### Model layer (primary accent)

- `indigo_velvet` (`r primary[["indigo_velvet"]]`) — main model
- `amethyst` (`r primary[["amethyst"]]`) — secondary model / comparison

### Highlight layer

- `orange` (`r primary[["orange"]]`) — emphasis, key results

### Live colour swatches

```{r}
#| label: colour-swatches
#| echo: false
#| fig-width: 8
#| fig-height: 1.8
#| fig-alt: "Colour swatches for the course brand palette"

swatch_df <-
  tibble::tibble(
    name = names(primary),
    hex  = unlist(primary),
    x    = seq_along(primary)
  ) |>
  dplyr::mutate(
    # Choose readable label colour based on luminance
    label_col = dplyr::if_else(
      name %in% c("graphite"),
      primary[["parchment"]],
      primary[["graphite"]]
    )
  )

ggplot2::ggplot(
  swatch_df,
  ggplot2::aes(x = x, y = 1)
) +
  ggplot2::geom_tile(
    ggplot2::aes(fill = hex),
    width = 0.95,
    height = 0.7
  ) +
  ggplot2::geom_text(
    ggplot2::aes(label = name, colour = label_col),
    y = 1.08,
    size = 3,
    fontface = "bold",
    vjust = 0
  ) +
  ggplot2::geom_text(
    ggplot2::aes(label = hex, colour = label_col),
    y = 0.92,
    size = 2.6,
    vjust = 1
  ) +
  ggplot2::scale_fill_identity() +
  ggplot2::scale_colour_identity() +
  ggplot2::scale_x_continuous(expand = ggplot2::expansion(mult = 0.05)) +
  ggplot2::scale_y_continuous(limits = c(0.6, 1.4)) +
  ggplot2::labs(x = NULL, y = NULL, title = "Course colour palette") +
  ggplot2::theme_void() +
  ggplot2::theme(
    plot.title = ggplot2::element_text(
      size = 11,
      face = "bold",
      colour = primary[["graphite"]],
      margin = ggplot2::margin(b = 6)
    ),
    plot.background = ggplot2::element_rect(
      fill = primary[["parchment"]],
      colour = NA
    )
  )
```

---

## Semantic Mapping (MANDATORY)

| Role              | Colour name    | Hex value                       | Usage                                  |
| ----------------- | -------------- | ------------------------------- | -------------------------------------- |
| Data              | `grey_olive`   | `r primary[["grey_olive"]]`     | raw observations, background structure |
| Model (primary)   | `indigo_velvet`| `r primary[["indigo_velvet"]]`  | main fitted curve, main inference      |
| Model (secondary) | `amethyst`     | `r primary[["amethyst"]]`       | alternative models, comparisons        |
| Highlight         | `orange`       | `r primary[["orange"]]`         | key insight, important value, warning  |

Rules:

- Grey is **always** data
- Purple family is **always** modelling
- Orange is used **sparingly** for emphasis

---

## Usage Rules

### General

- Use maximum **3 colours per figure**
- Maintain strong contrast hierarchy
- Prefer restraint over expressiveness

### Data (points)

- Colour: `grey_olive`
- Alpha: 0.2–0.4
- Size: small

### Models (lines)

- Primary model: `indigo_velvet`
- Secondary model: `amethyst`
- Line width: slightly thicker than default

### Highlights

Use `orange` only for:

- key datapoints
- annotated regions
- important labels

Do **NOT**:

- colour entire plots orange
- use orange for decorative purposes

### Uncertainty

- Use transparency of purple (`alpha = 0.15`)
- Or lighter amethyst
- Avoid introducing new colours

---

## Typography

Font definitions are in [`quarto/fonts.json`](quarto/fonts.json).

### Primary (sans-serif)

- Body: **`r fonts_data$body`** (weights 400, 500, 700)
- Headings: **`r fonts_data$heading`** (weights 400, 600, 700)

### Monospace

- **`r fonts_data$monospace`**; fallback: Fira Code

Usage:

- Monospace for code, formulas, filenames
- Optional: purple accent (`indigo_velvet`) for model formulas

---

## Layout Principles

- Left-aligned layouts
- Generous whitespace
- One main visual per slide

Optional structural motif:

- Thin vertical purple line for consistent framing

---

## Figure Design Pattern

Standard layer order:

1. Grey points (data)
2. Purple curve (model)
3. Optional secondary curve (amethyst)
4. Orange highlight (if needed)

This order reflects the analytical workflow (§2.1).

### Biological framing requirement

Every figure in a student-facing context must be anchored in a concrete biological or
ecological question.

Required:

- **Caption or slide narrative** states the biological question the figure addresses
  (e.g. *"Závisí délka křídla na druhu tučňáka?"*)
- **Axis labels** use ecologically meaningful Czech names with units — never raw column names
- **Annotations** (if orange) explain the biological significance, not just the statistical value

A figure is complete only when a student could answer "what biological question does this
address?" from the figure and its caption alone.

---

## ggplot2 Implementation

### Named colour constants

Define these at the top of every setup chunk and reuse throughout:

```r
clr_data      <- "#8A8A8A"   # grey_olive    — raw observations
clr_model     <- "#5D2890"   # indigo_velvet — primary model
clr_model2    <- "#86579E"   # amethyst      — secondary model / comparison
clr_highlight <- "#F3A712"   # orange        — emphasis / key result
clr_bg        <- "#F4F1EC"   # parchment     — HTML document background
```

### Standard course ggplot2 theme (`theme_biostat`)

Course figures use `theme_biostat()` by default. It is defined in each lesson's
generated `R/set_r_theme.R` and installed globally with `theme_set()`, so adding
it to every plot is unnecessary. A lesson may deliberately replace the complete
theme when students are learning to construct a figure themselves; such an
override is a pedagogical choice, not a brand synchronization failure.

The generated theme chooses its outer background by output format: parchment in
HTML, white in PDF and ordinary R use. The data panel remains white. This avoids
visible figure cards in PDF while preserving the parchment HTML surface without
depending on graphics-device transparency.

Reference implementation:

```r
theme_biostat <- function(base_size = 11, base_family = "Source Sans 3") {
  output_background <-
    if (knitr::is_html_output()) "#F4F1EC" else "#FFFFFF"

  ggplot2::theme_bw(base_size = base_size, base_family = base_family) +
    ggplot2::theme(
      plot.background   = ggplot2::element_rect(fill = output_background, colour = NA),
      panel.background  = ggplot2::element_rect(fill = "white", colour = NA),
      panel.grid.minor  = ggplot2::element_blank(),
      panel.border      = ggplot2::element_blank(),
      axis.line         = ggplot2::element_line(colour = "#2E2E2E", linewidth = 0.4),
      axis.text         = ggplot2::element_text(colour = "#2E2E2E"),
      axis.title        = ggplot2::element_text(colour = "#2E2E2E"),
      plot.title        = ggplot2::element_text(colour = "#2E2E2E", face = "bold"),
      legend.background = ggplot2::element_rect(fill = output_background, colour = NA),
      legend.key        = ggplot2::element_rect(fill = output_background, colour = NA),
      legend.position   = "bottom"
    )
}
```

### Example: scatter plot (data only) {.unnumbered}

```{r}
#| label: example-scatter-data
#| echo: true
#| fig-width: 6
#| fig-height: 4
#| fig-alt: "Example scatter plot showing grey data points on a parchment background"

# Synthetic data for illustration
set.seed(1)
n <- 80
example_df <-
  tibble::tibble(
    x = rnorm(n, mean = 20, sd = 5),
    y = 0.8 * x + rnorm(n, sd = 3)
  )

ggplot2::ggplot(
  example_df,
  ggplot2::aes(x = x, y = y)
) +
  ggplot2::geom_point(
    colour = clr_data,
    alpha  = 0.35,
    size   = 2
  ) +
  ggplot2::labs(
    x     = "Predictor variable (units)",
    y     = "Response variable (units)",
    title = "Data only — grey points on a white data panel"
  ) +
  theme_biostat()
```

### Example: scatter + fitted model {.unnumbered}

```{r}
#| label: example-scatter-model
#| echo: true
#| fig-width: 6
#| fig-height: 4
#| fig-alt: "Example scatter plot showing grey data points and a purple fitted line"

ggplot2::ggplot(
  example_df,
  ggplot2::aes(x = x, y = y)
) +
  ggplot2::geom_point(
    colour = clr_data,
    alpha  = 0.35,
    size   = 2
  ) +
  ggplot2::geom_smooth(
    method    = "lm",
    colour    = clr_model,
    linewidth = 1.2,
    se        = FALSE,
    formula   = y ~ x
  ) +
  ggplot2::labs(
    x     = "Predictor variable (units)",
    y     = "Response variable (units)",
    title = "Grey data and purple model"
  ) +
  theme_biostat()
```

---

## Consistency Rules

- Do not reassign colour meanings
- Do not introduce new hues without justification
- Maintain consistency across lectures and materials

---

## Guiding Principle

> Colour encodes epistemic status, not category.

- grey → observed
- purple → modelled
- orange → interpreted

---

## Default Decision Rules

Use the following defaults unless there is a strong reason to deviate:

1. Map data to grey (`grey_olive`)
2. Map the primary model to `indigo_velvet`
3. Use `amethyst` only for secondary comparison
4. Use `orange` only for deliberate emphasis
5. Limit each figure to the smallest viable palette
6. Prefer minimalism, consistency, and semantic clarity

Safe default (for most figures): grey points + purple model + no accent colour.

---

## Future Extensions

Possible controlled extensions:

- line types for additional models (instead of adding colours)
- transparency gradients for density
- annotations instead of additional colours

Avoid expanding the palette unless strictly necessary.

---

## Presentations (Quarto RevealJS)

### Slide background and theme

- Background colour: `parchment` (`r clr_bg`)
- Primary text: `graphite` (`r clr_text`)
- Code highlight style: `github-light` (consistent with parchment background)

### Typography on slides

- Headings: Source Sans 3 or Inter, bold
- Body: same sans-serif family at regular weight
- Code: JetBrains Mono or Fira Code
- Do NOT use more than one heading weight per slide

### Colour use in slide text and callouts

- Use `indigo_velvet` for highlighted model formulas or key statistical terms — apply sparingly
- Use `orange` for "question" labels, key-result callout borders, or "important" boxes
- Do NOT colour regular prose text for decoration
- Callout with left purple border = supplementary explanation; orange border = call to attention

### Figure density

- One figure per slide is the default
- Two-column layout is allowed for: side-by-side model comparisons, before/after reveals,
  data vs model overlays
- Never place more than two distinct figures on a single slide

### Presentation crafting contract

The template presentation contains a working gallery of the available
RevealJS components. The rules below are the canonical decision guide for
using that gallery in a real lesson.

- Build each concept as **data moment → observation or prediction → concept
  name → formalisation**. A statistical term should not open its own concept
  block.
- Give each slide one dominant message or teaching move.
- Alternate structured components with plain figures, code/output layouts,
  short questions, and minimal transitions.
- In any run of five consecutive slides, no more than three should use a named
  card, panel, box, blockquote, or strip as the primary visual element.
- Colour has semantic meaning: graphite or grey for observed data,
  indigo_velvet for models and estimates, and orange for interpretation,
  uncertainty, or warnings.
- Match highlighted words to the corresponding marks, lines, intervals, and
  component accents on the slide.

Learning outcomes are a stable exception to the general preference for more
visual slide structures. They should remain a concise, incremental **bullet
list**, not a grid of panels or cards. Use inline highlights inside the bullets
to make the statistical contrasts easy to scan:

```markdown
## Výsledky učení

Po dnešní přednášce budete schopni:

::: {.incremental}
- rozlišit [odhad efektu]{.text-hgl-indigo_velvet}
  od jeho [nejistoty]{.text-hgl-orange}
- interpretovat [95% interval spolehlivosti]{.text-hgl-orange}
- zapsat výsledek s [odhadem]{.text-hgl-indigo_velvet},
  [intervalem a jednotkami]{.text-hgl-orange}
:::
```

### Five slide families

A coherent lecture cycles through five functional slide families. Students
should be able to recognise what a slide is asking them to do before reading
every word.

|Family|Purpose|Typical treatment|
|---|---|---|
|Interaction|prediction, MCQ, think-pair-share, silent write|plain bold question or `.panel-question`|
|Data reveal|new graph, table, code output, or visual layer|full-width visual or code/output layout|
|Interpretation|read the evidence and state what it supports|graph plus `.panel-result`, `.strip-*`, or `.takeaway-strip`|
|Rule or warning|heuristic, misconception, limitation|`.card-rule`, `.card-warning`, or light warning panel|
|Transition or summary|reset attention, connect blocks, close the lesson|minimal prompt, short bullet summary, or compact blockquote|

No block of five or more slides should consist of only one family. Repetition
is justified when the same-data-one-new-layer sequence is itself the teaching
move, but the sequence should then end with a visibly different interpretation
or checkpoint slide.

### Component hierarchy

Choose the lightest component that communicates the intended structure.

|Family|Visual weight|Use for|Do not use for|
|---|---|---|---|
|Plain slide|none|full-width graph, short prompt, transition, learning outcomes|ordinary long explanation|
|`.panel-*`|light|question, result, data context, short comparison|turning every bullet into a separate tile|
|`.card-*`|heavy|rule, misconception, warning, major decision|routine conclusions or repeated prompts|
|`.strip-*`|light and horizontal|one closing statement, workflow step, semantic takeaway|multi-paragraph explanation or a second slide heading|
|`.takeaway-strip`|light and prominent|the main conclusion that closes a slide or short sequence|every interaction instruction|
|`.blockquote`|medium|short definition, biological frame, quoted principle|generic body prose|

The `.box-*` names are semantic aliases. Their visual weight is not uniform:

|Alias|Maps to|Weight|Primary use|
|---|---|---|---|
|`.box-question`|`.panel-question`|light|MCQ or reflection question|
|`.box-result`|`.panel-result`|light|main analytical result|
|`.box-data`|`.panel-data`|light|dataset or measurement context|
|`.box-rule`|`.card-rule`|heavy|rule or first-choice heuristic|
|`.box-warning`|`.card-warning`|heavy|methodological warning|

Panels use `[text]{.panel-title}` or `[text]{.panel-lead}` for their internal
heading. Cards use `[text]{.card-title}` as the first element so the coloured
top band is generated correctly.

### Panels and cards

Use panels when the information benefits from a visible category but should
remain part of the normal slide flow:

```markdown
::: {.panel-result}
[Co už víme]{.panel-title}

Směr vztahu je kladný a odhad sklonu je 4,05 g/mm.
:::
```

Use cards only when the content must carry more visual weight:

```markdown
::: {.card-warning}
[Pozor na interpretaci]{.card-title}

Viditelná asociace sama o sobě neprokazuje kauzalitu.
:::
```

Contrasting panels are appropriate when the contrast itself is the concept,
for example “co už víme” versus “co ještě nevíme”. Do not use a two-by-two panel
grid merely to make an ordinary list look more designed.

### Strips and takeaway strips

Strips are the lowest-weight named component. They keep a short conclusion
attached to the evidence above it without turning the entire slide into a
panel. Use no more than one dominant strip on a slide.

```markdown
::: {.strip-indigo}
Sklon říká, jak rychle se mění průměrná odezva s prediktorem.
:::

::: {.strip-orange}
Šířka intervalu vyjadřuje nejistotu odhadu.
:::

::: {.takeaway-strip}
**Hlavní závěr.** Uvádíme velikost efektu i její nejistotu.
:::
```

Semantic strip variants:

|Class|Meaning|
|---|---|
|`.strip-indigo`|model, estimate, or statistical result|
|`.strip-orange`|interpretation, uncertainty, or point requiring attention|
|`.strip-amethyst`|reasoning step, method, or rule|
|`.strip-olive`|data context or descriptive observation|
|`.strip-red`|error, invalid inference, or serious warning|
|`.takeaway-strip`|cross-layer conclusion using the brand gradient|

Add `.strip-compact` only when two closely related conclusions genuinely need
reduced vertical spacing. A strip should contain one sentence or one compact
list, not a paragraph of lecture notes.

### Inline highlights

Use `.text-hgl-*` primarily inside body text, bullets, questions, and takeaway
sentences. Highlighting only slide headings does not help students scan a
verbose slide.

```markdown
Naměřený bod je [pozorování]{.text-hgl-graphite}, fialový bod je
[odhad modelu]{.text-hgl-indigo_velvet} a vzdálenost mezi nimi je
[residuum]{.text-hgl-orange}.
```

Available highlight colours are `indigo_velvet`, `orange`, `amethyst`,
`grey_olive`, `graphite`, `white`, `light_gray`, and `parchment`. Prefer the
three semantic colours. Never assign a highlighted word a colour that
contradicts the figure or legend representing the same concept.

### Blockquotes, backgrounds, and layout utilities

Use `::: {.blockquote}` for one short definition, biological framing
statement, or quoted principle. Use the styled div rather than Markdown `>`
syntax. A blockquote is not a general-purpose text box.

Background classes are appropriate for true divider or reset slides:

|Background|Text treatment|Recommended use|
|---|---|---|
|`.bg-indigo`, `.bg-amethyst`, `.bg-olive`, `.bg-red`, `.bg-graphite`|white text|rare section divider or major reset|
|`.bg-orange`, `.bg-white`, `.bg-parchment`|graphite text|light divider, contrast slide, or example comparison|

Do not use background changes merely to make a routine content slide look
different. The colour must reinforce the function of the slide.

Use `::: {.columns .align-top}` when unequal columns should begin at the same
vertical position. Plain `.columns` may remain vertically centred when that
balance is intentional. The utility classes `.slide-margin-top-15`,
`.slide-margin-top-25`, and `.center-vertical` are available for isolated
elements, but normal document flow is preferred whenever possible.

### Roughnotation

The `quarto-roughnotation` shorthand is a live-teaching accent, not a second
component system:

```markdown
[klíčová fráze]{.rn-underline-indigo}
[0,76]{.rn-circle-orange}
[část vzorce]{.rn-bracket-indigo}
```

Use at most one roughnotation mark on a typical slide and only a few across a
lecture. It may underline one phrase, circle one result, or bracket one part of
an equation after students have had time to look. Do not annotate paragraphs,
decorate all answer choices, or combine several roughnotation colours on one
slide. A semantic text highlight or direct plot annotation is the default;
roughnotation is reserved for a deliberate live gesture.

### Progressive disclosure

- Reveal layers in this order: data (grey) → model (purple) → interpretation (orange callout)
- Use RevealJS fragments or incremental list options
- Never show a model result before posing a prediction question to the audience
- Use `.incremental` for learning outcomes and lists in which each item is a
  new idea.
- Wrap a complete panel, card, or strip in `.fragment` when the whole component
  should appear after discussion.
- Progressive disclosure should reveal reasoning, not merely slow down a list
  that students could read more clearly at once.

### Interaction slides

- Make interaction slides recognisable, but vary their furniture across the
  lecture: voting cards, a light question panel, a silent-write prompt, or an
  unframed centred question can all be appropriate.
- Use the same visual treatment for all answer options within one question.
- Reveal the correct answer only after students commit; add one
  brand-consistent cue rather than restyling the whole option.
- The word "Otázka" or "Diskuze" may be accented in orange, but the task itself
  must remain the dominant element.
- Follow a prediction with a result and a visible interpretation or
  checkpoint.

### Code display on slides

- Keep visible code chunks to ≤ 8 lines; scrolling is acceptable for demos
- Code output inherits the parchment background — do not override with white or grey
- Student-visible code uses base R only (no tidyverse prefixes)

---

## Skripta (Learning Materials — Quarto HTML / Typst PDF)

### HTML theme

- Base Quarto theme: `flatly` + custom `skripta_theme.scss` (generated from JSON)
- TOC: always shown (`toc: true`, `toc-depth: 3`), left sidebar in HTML
- Section numbering: always enabled (`number-sections: true`)

### Callout blocks

Use Quarto callouts for exactly four purposes:

| Callout type | Purpose | Accent colour |
| ------------ | ------- | ------------- |
| `note` | Supplementary explanation, extra detail | indigo_velvet (SCSS customised) |
| `tip` | Optional technical context that is not needed for the main argument | grey_olive |
| `important` | Key concept or definition the student must retain | indigo_velvet |
| `warning` | Common mistake or statistical misconception | orange |

Use `tip` only with the Czech heading **Doplňující** (optionally followed by a short descriptor). For a long optional code example, place a brief `tip` callout before the code and fold the code in HTML; the code remains in the normal page flow in PDF. A short self-contained technical example may instead be placed directly in a collapsible `tip` callout, which remains expanded in PDF. Avoid `caution` and do not invent further callout types.

### Colour in prose

- Do not colour prose text
- **Bold** is the only allowed inline emphasis for key terms appearing for the first time
- *Italic* is reserved for species names and foreign-language terms — not for general emphasis
- Code inline spans are used for all R objects, function names, variable names, and file paths

### Figures in skripta

- Semantic colour mapping applies identically to skripta figures
- Every figure requires a caption that **interprets** the figure — not just describes it
- Default width: `out.width = "80%"` for HTML; `out.width = "100%"` for Typst/PDF
- Always include `fig-alt` alt text for accessibility

### Markdown tables

Markdown tables use one shared treatment across skripta HTML, Typst/PDF, and
RevealJS:

- full available width, with wrapping inside cells;
- indigo_velvet header with white bold text;
- visible but quiet borders between cells;
- sufficient cell padding for scanning;
- alternating white and parchment/light-gray body rows.

These rules are generated from `quarto/custom_theme.json`. Do not restyle an
individual lesson table to compensate for unformatted PDF defaults; correct the
shared generator and visually verify representative HTML and Typst/PDF tables.

### Interactive HTML components

#### Glossary tooltips

- Tooltip background: `#FFFFFF` (white) — must contrast against the parchment (`#F4F1EC`) page background
- Border: `1px solid rgba(93, 40, 144, 0.25)` — subtle indigo_velvet tint
- Shadow: `0 4px 12px rgba(0, 0, 0, 0.12)` — lifts the tooltip visually off the page
- The CSS is emitted by `nastaveni-slovnik` chunk; the `glossary_style()` call must use `def_bg = "#FFFFFF"`

#### Tabsets (`.panel-tabset`)

Tab button states use existing colour tokens to give clear click affordance against the parchment background:

| State | Background | Text | Border |
| ----- | ---------- | ---- | ------ |
| Inactive | `light_gray` (`#F0EDE8`) | `graphite` | `rgba(0,0,0,0.12)` |
| Hover | `amethyst` (`#86579E`) | white | `amethyst` |
| Active | `indigo_velvet` (`#5D2890`) | white | `indigo_velvet` |

This styling is generated automatically by `generate_skripta_html_theme.R` and must not be overridden manually.

### Code blocks

- Student-visible chunks (`echo: true`): base R, no prefixes for preloaded packages
- Hidden setup chunks (`echo: false`): tidyverse with full namespace (`pkg::fn()`)
- Code folding is acceptable only for lengthy raw data preparation

---

## R Visualisations — Extended Reference

### Model uncertainty bands

Show uncertainty via transparent fill using the model colour — do NOT introduce a new colour:

```r
ggplot2::geom_ribbon(
  ggplot2::aes(ymin = lower, ymax = upper),
  fill  = clr_model,
  alpha = 0.15
)
```

### Two-model comparison

Use a thinner, dashed line for the secondary model to reinforce the visual hierarchy:

```r
# Primary model
ggplot2::geom_smooth(colour = clr_model,  linewidth = 1.2, se = FALSE)
# Secondary model
ggplot2::geom_smooth(colour = clr_model2, linewidth = 1.0, se = FALSE, linetype = "dashed")
```

### Boxplots and violin plots

Data points underneath a transparent model-coloured box:

```r
ggplot2::geom_jitter(colour = clr_data,  alpha = 0.3,  width = 0.15, size = 1.5) +
ggplot2::geom_boxplot(colour = clr_model, fill = NA, linewidth = 0.8, outlier.shape = NA)
```

### Categorical group colour

When colour **must** encode a grouping variable (e.g. species, treatment):

- Use a distinct palette that does NOT overlap with the semantic palette
- Never reuse `clr_model` or `clr_highlight` as a group colour
- Document the override with a comment
- Suggested accessible palette for up to 4 groups:

```r
clr_skupiny <- c(
  "skupina_A" = "#4E9A8A",   # teal
  "skupina_B" = "#C2613A",   # terracotta
  "skupina_C" = "#2E6CA4",   # steel blue
  "skupina_D" = "#7D7D7D"    # dark grey
)
```

### Highlights and annotations

```r
# Highlighted individual point
ggplot2::geom_point(
  data   = data_highlighted,
  colour = clr_highlight,
  size   = 4
)

# Annotation label
ggplot2::annotate(
  geom     = "text",
  x        = x_pos,
  y        = y_pos,
  label    = "Klíčový výsledek",
  colour   = clr_highlight,
  fontface = "bold",
  size     = 4
)
```

### Figure dimensions by output context

| Context | `fig.width` | `fig.height` | `out.width` |
| ------- | ----------- | ------------ | ----------- |
| Presentation slide | 7 | 5 | `"100%"` |
| Skripta HTML | 7 | 5 | `"80%"` |
| Skripta PDF (Typst) | 6 | 4 | `"100%"` |

### Axis labels and Czech language

- Always use Czech axis labels in student-visible plots; set with `ggplot2::labs()`
- Never leave default English column-name axis labels in student-facing output
- Include units in parentheses where relevant: `"Délka křídla (mm)"`
- Annotation text passed to `ggplot2::annotate()` must be Czech
- Legend labels and guide titles must be Czech

See §1.1 for the global Czech language requirement.

---

## GitHub / Repository Identity

### README files

- Language: English (technical documentation is public-facing)
- Content: state what the repo contains, how to render locally, and who the audience is
- Keep concise; do not include personal context or course history

### Organisation profile (`.github` repo)

- The `README.md` in the `.github` repo is the public face of the organisation
- Colour references should use hex values where relevant
- Tone: informative and neutral

### Release assets and file naming

- Rendered outputs use predictable, fixed names: `presentation.html`, `skripta.html`, `skripta_raw.pdf`
- File names must not contain spaces or special characters (including Czech diacritics)
- Do not change naming conventions without updating the corresponding GitHub Actions workflow
